---
title: "Data Cleaning and Exploratory Analysis"
author: "Annabella Hines"
date: "2023-02-24"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(tidyverse)
```


## Load the data.

```{r}
#Path to data.

life_data_location <- here::here("data","processed_data","processeddata_life.rds")

risk_data_location <- here::here("data","processed_data","processeddata_risk.rds")

#load data

life <- readRDS(life_data_location)

risk <- readRDS(risk_data_location)
```


# Data exploration through tables

First we'll use some code to produce and save a summary table for each data set.

```{r}
summary_lifeexp <- skimr::skim(life)

print(summary_lifeexp)

summary_riskfactor <- skimr::skim(risk)

print(summary_riskfactor)

# save to file

summarytable_lifeexp = here("results", "summarytablelifeexp.rds")

saveRDS(summary_lifeexp, file = summarytable_lifeexp)

summarytable_riskfactor = here("results", "summarytableriskfactor.rds")

saveRDS(summary_riskfactor, file = summarytable_riskfactor)


skimr::skim(life)
lifedes<-psych::describe(life)
lifedesrds = here("results", "lifedes.rds")
saveRDS(lifedes, file=lifedesrds)

riskdes <- psych::describe(risk)
riskdesrds = here("results", "riskdes.rds")
saveRDS(riskdes, file=riskdesrds)

```


# Data exploration through figures

First a quick look at Adult mortality and infant mortality plotted against life expectancy while also representing development status.
```{r}
##Plot of adult mortality vs life expectancy with point size correlated to infant deaths and color to status of country
ggplot(life, aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()
#Same variables but for the year 2000
p1_2000<- filter(life, Year==2000) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()+ylab("Life Expectancy")+ xlab("Adult Mortality")+labs(title="Life Expectancy vs Adult Mortality with Infant Deaths and Country Status 2000")
#Save plot.
plot1 = here("results","p1_2000.png")
ggsave(filename = plot1, plot=p1_2000) 
#Same variables but for 2015
p2_2015<-filter(life, Year==2015) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, color=Status, size=Infant.deaths))+geom_point()+geom_jitter()+ylab("Life Expectancy")+ xlab("Adult Mortality")+labs(title="Life Expectancy vs Adult Mortality with Infant Deaths and Country Status 2015")
#save plot.
plot2 = here("results","p2_2015.png")
ggsave(filename = plot2, plot=p2_2015)
```
The trends appear to have stayed fairly consistent throughout the period from 2000 to 2015 with a higher life expectancy associated with lower adult and infant mortality and being classified as a developed nation.

```{r}
#comparing life expectancy to GDP in 2015
combo_both <- right_join(life, risk, by=c("Country", "Year")) %>% filter(Year==c(2000, 2015)) %>% drop_na(Status)
combo_2015 <- right_join(life, risk, by=c("Country", "Year")) %>% filter(Year==2015)
combo_2000 <- right_join(life, risk, by=c("Country", "Year")) %>% filter(Year==2000)
ggplot(combo_2015, aes(x=GDP, y=Life.expectancy, color=Status))+geom_point()
ggplot(combo_2000, aes(x=GDP, y=Life.expectancy, color=Status))+geom_point()
```
Here is life expectancy against GDP in 2015, which has a lot of variation in life expectancy when GDP is less that 10000. Once past that threshold, the life expectancies remain in the high range.

Next I wanted to look at some vaccination rates in comparison to life expectancy.
```{r}
##Hepatitis.B 2015
ggplot(combo_2015, aes(y=Life.expectancy, x=Hepatitis.B))+geom_point()+geom_smooth()
##Hepatitis.B 2000
ggplot(combo_2000, aes(y=Life.expectancy, x=Hepatitis.B))+geom_point()+geom_smooth()
#Polio 2015
ggplot(combo_2015, aes(y=Life.expectancy, x=Polio))+geom_point()+geom_smooth()
#Polio 2000
ggplot(combo_2000, aes(y=Life.expectancy, x=Polio))+geom_point()+geom_smooth()
#Diptheria 2015
ggplot(combo_2015, aes(y=Life.expectancy, x=Diphtheria))+geom_point()+geom_smooth()
#Diptheria 2000
ggplot(combo_2000, aes(y=Life.expectancy, x=Diphtheria))+geom_point()+geom_smooth()
```
The vaccination rates are difficult to visualize a trend for as most countries seem to have high vaccination rates, so I added trend lines. They all were very similar in shape with a rise in life expectancy at rates over 85 percent coverage.

```{r}
ggplot(life, aes(x=Status, y=Polio))+geom_boxplot()
```
Taking Polio vaccination rates as an example, developed countries have a much smaller distribution with higher coverage than developing countries that are much more varied.

```{r}
p1_2000 <- life %>% filter(Year==2000) %>% ggplot(aes(x=Status, y=Life.expectancy, fill=Status)) + geom_boxplot()+ylab("Life Expectancy")+xlab("Development Status")+labs(title="Life Expectancy Distribution by Development Status in 2000")
p1_2015 <- life %>% filter(Year==2015) %>% ggplot(aes(x=Status, y=Life.expectancy, fill=Status)) + geom_boxplot()+ylab("Life Expectancy")+xlab("Development Status")+labs(title="Life Expectancy Distribution by Development Status in 2015")

plot(p1_2000)
plot(p1_2015)

figure_file = here("results","LifeExpectancyStatusBoxPlot2000.png")

ggsave(filename = figure_file, plot=p1_2000) 

figure_file2 = here("results","LifeExpectancyStatusBoxPlot2015.png")

ggsave(filename = figure_file2, plot=p1_2015) 

##combine both plots with facet_wrap
sc <- ggplot(combo_both, aes(Status, Life.expectancy, fill=Status)) +geom_boxplot()+facet_wrap(vars(Year))

figure_file3 = here("results","LifeExpectancyStatusBoxPlotFull.png")

ggsave(filename = figure_file3, plot=sc) 
```
Developed countries have an on average higher life expectancy distribution than developing countries.

```{r}
riskUS <- filter(risk, Country=="United States")
p3 <- ggplot(riskUS, aes(x=Year))+geom_point(aes(y=Unsafe.sanitation), color="blue")+geom_point(aes(y=Unsafe.water.source),color="darkgreen")+labs(title="Deaths Due to Unsafe Water Source, Unsafe Sanitation, or No Access to Handwashing Facilities in the US")+geom_point(aes(y=No.access.to.handwashing.facility), color="red")+ylab("Deaths")
plot(p3)

figure_file4 = here("results","WaterriskUS.png")

ggsave(filename = figure_file, plot=p3) 
```


