---
title: "Data Cleaning and Exploratory Analysis"
author: "Annabella Hines"
date: "2023-02-24"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```


## Load the data.

```{r}
#Path to data.
life_data_location <- here::here("data","processed_data","processeddata_life.rds")
risk_data_location <- here::here("data","processed_data","processeddata_risk.rds")
#load data
life <- readRDS(life_data_location)
risk <- readRDS(risk_data_location)
```


# Data exploration through tables

First we'll use some code to produce and save a summary table for each data set.

```{r}
summary_lifeexp <- skimr::skim(life)
print(summary_lifeexp)
summary_riskfactor <- skimr::skim(risk)
print(summary_riskfactor)
# save to file
summarytable_lifeexp = here("results", "summarytablelifeexp.rds")
saveRDS(summary_lifeexp, file = summarytable_lifeexp)
summarytable_riskfactor = here("results", "summarytableriskfactor.rds")
saveRDS(summary_riskfactor, file = summarytable_riskfactor)
```


# Data exploration through figures

First a quick look at Adult mortality and infant mortality plotted against life expectancy.
```{r}
##Plot of adult mortality vs life expectancy with point size correlated to infant deaths and color to status of country
ggplot(life, aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()
#Same variables but for the year 2000
filter(life, Year==2000) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()
#Same variables but for 2015
filter(life, Year==2015) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, color=Status, size=Infant.deaths))+geom_point()+geom_jitter()
```
The trends appear to have stayed fairly consistent throughout the period with a higher life expectancy associated with lower adult and infant mortality and being classified as a developed nation.

```{r}
combo <- right_join(life, risk, by=c("Country", "Year"))
filter(combo, Year==c(2000, 2015)) %>% ggplot(aes(x=Air.pollution, y=Life.expectancy, color=Year))+geom_point()

```
```{r}
ggplot(life, aes(x=Year, y=Polio, color=Status))+geom_point()
```


```{r}
p1 <- life %>% filter(Year==2000) %>% ggplot(aes(x=Status, y=Life.expectancy)) + geom_boxplot() 
plot(p1)
figure_file = here("results","")
ggsave(filename = figure_file, plot=p1) 
```
Developed countries have an on average higher life expectancy distribution than developing countries.


```{r}
p2 <- mydata %>% ggplot(aes 
plot(p2)
figure_file = here("results",".png")
ggsave(filename = figure_file, plot=p2) 
```


```



# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.


