---
title: "Data Cleaning and Exploratory Analysis"
author: "Annabella Hines"
date: "2023-02-24"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here)
library(dplyr)
library(skimr)
library(ggplot2)
library(tidymodels)
library(tidyverse)
```


## Load the data.

```{r}
#Path to data.

life_data_location <- here::here("data","processed_data","processeddata_life.rds")

risk_data_location <- here::here("data","processed_data","processeddata_risk.rds")

#load data

life <- readRDS(life_data_location)

risk <- readRDS(risk_data_location)
```


# Data exploration through tables

First we'll use some code to produce and save a summary table for each data set.

```{r}
summary_lifeexp <- skimr::skim(life)

print(summary_lifeexp)

summary_riskfactor <- skimr::skim(risk)

print(summary_riskfactor)

# save to file

summarytable_lifeexp = here("results", "summarytablelifeexp.rds")

saveRDS(summary_lifeexp, file = summarytable_lifeexp)

summarytable_riskfactor = here("results", "summarytableriskfactor.rds")

saveRDS(summary_riskfactor, file = summarytable_riskfactor)

```


# Data exploration through figures

First a quick look at Adult mortality and infant mortality plotted against life expectancy.
```{r}
##Plot of adult mortality vs life expectancy with point size correlated to infant deaths and color to status of country
ggplot(life, aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()
#Same variables but for the year 2000
p1_2000<- filter(life, Year==2000) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, size=Infant.deaths, color=Status))+geom_point()+geom_jitter()+ylab("Life Expectancy")+ xlab("Adult Mortality")+labs(title="Life Expectancy vs Adult Mortality with Infant Deaths and Country Status 2000")
ggsave("p1_2000.png")
#Same variables but for 2015
p2_2015<-filter(life, Year==2015) %>% ggplot(aes(x=Adult.Mortality, y=Life.expectancy, color=Status, size=Infant.deaths))+geom_point()+geom_jitter()+ylab("Life Expectancy")+ xlab("Adult Mortality")+labs(title="Life Expectancy vs Adult Mortality with Infant Deaths and Country Status 2015")
#save plots
ggsave("p2_2015.png")


```
The trends appear to have stayed fairly consistent throughout the period with a higher life expectancy associated with lower adult and infant mortality and being classified as a developed nation.

```{r}
combo <- right_join(life, risk, by=c("Country", "Year"))
filter(combo, Year==c(2000, 2015)) %>% ggplot(aes(x=Air.pollution, y=Life.expectancy, color=Year))+geom_point()

```
I need to rework this one
```{r}
filter(life, Year==2015) %>% select(Under.five.deaths, Country)
filter(life, Year==2015) %>% select(Under.five.deaths, Country) %>% top_n(Under.five.deaths, n=5)
filter(life, Year==2015) %>% select(Under.five.deaths, Country) %>% top_n(Under.five.deaths, n=-5)
filter(life, Year==2015) %>% filter(Country%in%c("Democratic Republic of the Congo", "Ethiopia", "India", "Nigeria", "Pakistan", "Austria", "Albania", "Belize" )) %>% ggplot(aes(x=Country, y=Under.five.deaths, color=Status))+geom_point()
```


```{r}
ggplot(life, aes(x=Status, y=Polio))+geom_boxplot()
```


```{r}
p1 <- life %>% filter(Year==2000) %>% ggplot(aes(x=Status, y=Life.expectancy)) + geom_boxplot() 

plot(p1)

figure_file = here("results","LifeExpectancyStatusBoxPlot.png")

ggsave(filename = figure_file, plot=p1) 
```
Developed countries have an on average higher life expectancy distribution than developing countries.

```{r}
p2 <- ggplot(riskUS, aes(x=Year))+geom_point(aes(y=Unsafe.sanitation), color="blue")+geom_point(aes(y=Unsafe.water.source),color="darkgreen")+ylab("Deaths Due to Unsafe Water Source, Unsafe Sanitation, or No Access to Handwashing Facilities in the US")+geom_point(aes(y=No.access.to.handwashing.facility), color="red")
plot(p2)

figure_file = here("results","WaterriskUS.png")

ggsave(filename = figure_file, plot=p2) 
```


```{r}
lm_mod <- linear_reg() %>% set_engine("lm")
fitlm1 <- lm_mod %>%
  fit(Life.expectancy ~ Year, data=life)
#checking results
fitlm1
tidy(fitlm1)
```
Year seemed to be positively correlated with life expectancy.

```{r}
life$Status <- as.factor(life$Status)
glimpse(life)
glm_mod <- logistic_reg() %>% set_engine("glm")
glm1 <- glm_mod %>% fit(Status ~ Life.expectancy, data=life)
#checking results
tidy(glm1)
```

```{r}
fitlm2 <- lm_mod %>%
  fit(Life.expectancy ~ BMI, data=life)
#checking results
fitlm2
tidy(fitlm2)
```


```{r}
fitlm3 <- lm_mod %>% fit(Life.expectancy~., data=life)
fitlm3
tidy(fitlm3)
```

```{r}
fitlm4 <- lm_mod %>% fit(Child.wasting~Non.exclusive.breastfeeding+Discontinued.breastfeeding, data=risk)
fitlm4
tidy(fitlm4)
```

